\textbf{Estimating the proportion of beneficial mutations that are not adaptive in mammals}

### 2024-06-19 manuscript/supp-mat.tex
Prepare for PLOS G. revisions

\usepackage[margin=50pt]{geometry}
\usepackage{amssymb,amsfonts,amsmath,amsthm,mathtools}
\usepackage{lmodern}
\usepackage{bm,bbold}
\usepackage{verbatim}
\usepackage{float}
\usepackage{listings, enumerate, enumitem}
\usepackage[export]{adjustbox}
\usepackage{tabu}
\usepackage{longtable}
\tabulinesep=0.6mm
\newcommand\cellwidth{\TX@col@width}
\usepackage{hhline}
\setlength{\arrayrulewidth}{1.2pt}
\usepackage{multicol,multirow,array}
\usepackage{etoolbox}
\AtBeginEnvironment{tabu}{\footnotesize}
\usepackage{booktabs}
\usepackage{makecell}
\usepackage{orcidlink}
\usepackage{graphicx}
\pdfinclusioncopyfonts=1

\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\thefigure}{S\arabic{figure}}

\newcommand{\UniDimArray}[1]{\bm{#1}}
\newcommand{\der}{\text{d}}
\newcommand{\e}{\text{e}}
\newcommand{\Ne}{N_{\text{e}}}
\newcommand{\proba}{\mathbb{P}}
\newcommand{\pfix}{\proba_{\text{fix}}}
\newcommand{\dn}{d_N}
\newcommand{\ds}{d_S}
\newcommand{\dnds}{\dn / \ds}
\newcommand{\Sphy}{S_{0}}
\newcommand{\SphyMean}{\overline{\Sphy}}
\newcommand{\SphyDel}{\mathcal{D}_0}
\newcommand{\SphyNeu}{\mathcal{N}_0}
\newcommand{\SphyBen}{\mathcal{B}_0}
\newcommand{\Sphyclass}{x}
\newcommand{\SphyclassAlt}{y}
\newcommand{\given}{\mid}
\newcommand{\Spop}{S}
\newcommand{\SpopDel}{\mathcal{D}}
\newcommand{\SpopNeu}{\mathcal{N}}
\newcommand{\SpopBen}{\mathcal{B}}
\newcommand{\ProbaPopDel}{\proba [ \SpopDel]}
\newcommand{\ProbaPopNeu}{\proba [ \SpopNeu ]}
\newcommand{\ProbaPopBen}{\proba [ \SpopBen ]}
\newcommand{\AdvMean}{\beta_b}
\newcommand{\DelMean}{\beta_d}
\newcommand{\thetaSyn}{\theta_{\text{S}}}

\title{\textbf{Mammalian protein-coding genes exhibit widespread beneficial mutations that are not adaptive}}

@@

\begin{document}
    \maketitle
    \part*{Supplementary materials}
    \tableofcontents
    \clearpage

@@

    \subsection{Mutation-selection codon models}
    \label{sec:site-specific-mutation-selection-codon-models}
    In \textit{BayesCode} (\url{https://github.com/ThibaultLatrille/bayescode}), we ran the mutation-selection codon models \textit{mutselomega} for 2000 points of MCMC with the options:
    \begin{scriptsize}
        \begin{verbatim}
 mutselomega ---omegashift 0.0 --ncat 30 -a my_alignment.phy -t my_tree.newick -u 2000 my_genename
        \end{verbatim}
    \end{scriptsize}
    The collection of site-specific fitness profiles ($\UniDimArray{F^{(i)}}, \forall i$) are then obtained by running \textit{readmutselomega}, reading 1000 points of MCMC (first 1000 are considered as burn-in) with the options:
    \begin{scriptsize}
        \begin{verbatim}
 readmutselomega --every 1 --until 2000 --burnin 1000 --ss my_genename
        \end{verbatim}
    \end{scriptsize}
    The gene-specific $4 \times 4$ nucleotide mutation rate matrix ($\UniDimArray{\mu}$) is also obtained by running \textit{readmutselomega}, reading 1000 points of MCMC (first 1000 are considered as burn-in) with the options:
    \begin{scriptsize}
        \begin{verbatim}
 readmutselomega --every 1 --until 2000 --burnin 1000 --nuc my_genename
        \end{verbatim}
    \end{scriptsize}

    \subsection{Summary tables}\label{subsec:summary-table-mutsel}

    \subsubsection{$\dnds$ for deleterious ($\SphyDel$), nearly-neutral ($\SphyNeu$) and beneficial back-mutations ($\SphyBen$)}
    \begin{center}
        \captionof{table}{$\dnds$ for deleterious ($\SphyDel$), nearly-neutral ($\SphyNeu$) and beneficial back-mutations ($\SphyBen$).}
        \scriptsize
        \begin{longtable*}{|l|l|r|r|r|r|r|r|}
            \toprule
            Population &             Species & Watterson's $\thetaSyn$  & $\proba_{div}[ \SphyBen ]$ & $\dnds $ & $\dn ( \SphyDel ) / \ds$ & $\dn ( \SphyNeu ) / \ds$ & $\dn ( \SphyBen ) / \ds$ \\
            \midrule
            \endhead
            \midrule
            \multicolumn{8}{r}{{Continued on next page}} \\
            \midrule
            \endfoot

@@

        \item Watterson's $\thetaSyn$ is the observed genetic diversity calculated for synonymous changes.

@@

        \item $\dn(\SphyDel) / \ds$ (eq.~6) is the ratio of non-synonymous over synonymous substitutions, when restricted to non-synonymous substitutions in the terminal branch that are $\SphyDel$.
        \item $\dn(\SphyNeu) / \ds$ (eq.~6) is the ratio of non-synonymous over synonymous substitutions, when restricted to non-synonymous substitutions in the terminal branch that are $\SphyNeu$.
        \item $\dn(\SphyBen) / \ds$ (eq.~6) is the ratio of non-synonymous over synonymous substitutions, when restricted to non-synonymous substitutions in the terminal branch that are $\SphyBen$.

@@

    \subsubsection{$\dnds$ over-estimation due to beneficial back-mutations}
    \begin{center}
        \captionof{table}{$\dnds$ over-estimation due to beneficial back-mutations.}
        \scriptsize
        \begin{longtable*}{|l|l|r|r|r|r|r|}
            \toprule
            Population & Species & Watterson's $\thetaSyn$ & $\proba_{div}[\Sphy > 1]$ & $\dnds $ & $\dn(\Sphy < 1) / \ds$ & $\delta(\dnds )$ \\
            \midrule
            \endhead
            \midrule
            \multicolumn{7}{r}{{Continued on next page}} \\
            \midrule
            \endfoot

>>>>

\usepackage{bbm}
\usepackage[margin=50pt]{geometry}
\usepackage{amssymb,amsfonts,amsmath,amsthm,mathtools}
\usepackage{lmodern}
\usepackage{bm,bbold}
\usepackage{verbatim}
\usepackage{float}
\usepackage{listings, enumerate, enumitem}
\usepackage[export]{adjustbox}
\usepackage{tabu}
\usepackage{longtable}
\tabulinesep=0.6mm
\newcommand\cellwidth{\TX@col@width}
\usepackage{hhline}
\setlength{\arrayrulewidth}{1.2pt}
\usepackage{multicol,multirow,array}
\usepackage{etoolbox}
\AtBeginEnvironment{tabu}{\footnotesize}
\usepackage{booktabs}
\usepackage{makecell}
\usepackage{orcidlink}
\usepackage{graphicx}
\usepackage{blkarray}
\usepackage{pgf,tikz}
\usetikzlibrary{shapes,arrows,backgrounds,fit,positioning,arrows,automata,calc}
\tikzset{res/.style={ellipse,draw,minimum height=1.0cm,minimum width=0.8cm}}
\tikzset{literal/.style={rectangle,draw,minimum height=0.5cm,minimum width=0.8cm,text width = 1.2 cm, align = center}}

\pdfinclusioncopyfonts=1

\renewcommand{\baselinestretch}{1.5}
\renewcommand{\arraystretch}{0.6}
\frenchspacing

\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\theequation}{S.\arabic{equation}}

\newcommand{\UniDimArray}[1]{\bm{#1}}
\newcommand{\BiDimArray}[1]{\bm{#1}}

\newcommand{\der}{\text{d}}
\newcommand{\e}{\text{e}}
\newcommand{\Ne}{N_{\text{e}}}
\newcommand{\proba}{\mathbb{P}}
\newcommand{\pfix}{\proba_{\text{fix}}}
\newcommand{\dn}{d_N}
\newcommand{\ds}{d_S}
\newcommand{\dnds}{\dn / \ds}
\newcommand{\Sphy}{S_{0}}
\newcommand{\SphyMean}{\overline{\Sphy}}
\newcommand{\SphyDel}{\mathcal{D}_0}
\newcommand{\SphyNeu}{\mathcal{N}_0}
\newcommand{\SphyBen}{\mathcal{B}_0}
\newcommand{\Sphyclass}{x}
\newcommand{\SphyclassAlt}{y}
\newcommand{\given}{\mid}
\newcommand{\Spop}{S}
\newcommand{\SpopDel}{\mathcal{D}}
\newcommand{\SpopNeu}{\mathcal{N}}
\newcommand{\SpopBen}{\mathcal{B}}
\newcommand{\ProbaPopDel}{\proba [ \SpopDel]}
\newcommand{\ProbaPopNeu}{\proba [ \SpopNeu ]}
\newcommand{\ProbaPopBen}{\proba [ \SpopBen ]}
\newcommand{\AdvMean}{\beta_b}
\newcommand{\DelMean}{\beta_d}
\newcommand{\thetaSyn}{\theta_{\text{S}}}
\newcommand{\pvalue}{p\text{-value}}

% Model
\newcommand{\submatrix}{q}
\newcommand{\Submatrix}{\BiDimArray{\submatrix}}
\newcommand{\probmatrix}{P}
\newcommand{\Probmatrix}{\BiDimArray{\probmatrix}}
\newcommand{\fit}{F}
\newcommand{\Fit}{\UniDimArray{\fit}}
\newcommand{\indice}{l}
\newcommand{\indiceexp}{^{(\indice)}}
\newcommand{\ci}{{a}}
\newcommand{\cj}{{b}}
\newcommand{\itoj}{\ci \mapsto \cj}
\newcommand{\nuc}{\mathcal{M}}
\newcommand{\fiti}{\fit_{\ci}}
\newcommand{\fitj}{\fit_{\cj}}
\newcommand{\mutmatrix}{R}
\newcommand{\Mutmatrix}{\BiDimArray{\mutmatrix}}
\newcommand{\exchan}{\rho}
\newcommand{\Exchan}{\UniDimArray{\exchan}}
\newcommand{\mutequi}{\sigma}
\newcommand{\Mutequi}{\UniDimArray{\mutequi}}
\newcommand{\Tree}{\mathcal{T}}
\newcommand{\branch}{\text{j}}
\newcommand{\branchexp}{^{(\branch)}}
\newcommand{\branchlength}{l}
% Alignment
\newcommand{\data}{D}
\newcommand{\Data}{\BiDimArray{\data}}
\newcommand{\site}{\text{i}}
\newcommand{\Nsite}{\text{N}}
\newcommand{\siteexp}{^{(\site)}}
\newcommand{\Setsite}{\site \in \{1, \hdots, \Nsite\} }
\newcommand{\branchsiteexp}{^{(\branch, \site)}}
% Categories
\newcommand{\cat}{\text{k}}
\newcommand{\Ncat}{\text{K}}
\newcommand{\catexp}{^{(\cat)}}
\newcommand{\catInterval}{\{1, \hdots, \Ncat\}}
\newcommand{\Setcat}{\cat \in \catInterval }
\newcommand{\branchcatexp}{^{(\branch, \cat)}}
\newcommand{\profile}{\phi}
\newcommand{\Profile}{\UniDimArray{\profile}}
\newcommand{\concentrationProfile}{\alpha}
\newcommand{\centerProfile}{\UniDimArray{\gamma}}
\newcommand{\catVar}{\kappa}
\newcommand{\catsite}{\catVar\left(\site\right)}
\newcommand{\catmultivar}{m}
\newcommand{\catMultiVar}{\UniDimArray{\catmultivar}}
\newcommand{\stickbreaking}{\theta}
\newcommand{\StickBreaking}{\UniDimArray{\stickbreaking}}
\newcommand{\stick}{\psi}
\newcommand{\stickbreakinghyper}{\beta}
\newcommand{\Multivariate}{\UniDimArray{Z}}
\newcommand{\subhistory}{\mathcal{H}}

\title{\textbf{Estimating the proportion of beneficial mutations that are not adaptive in mammals}}

@@

\begin{document}
    \maketitle
    \part*{Supplementary materials}
    \tableofcontents

    \newpage
    \listoffigures

    \newpage
    \listoftables

    \clearpage
    \section{Mutation-selection codon models}\label{sec:mutsel}

    \subsection{Prior distributions and parameters of the model}\label{subsec:priors}

    The parameterization of the models is described as a Bayesian hierarchical model, including the prior distributions and the parameters of the model.
    This hierarchical model is formally represented as directed acyclic graph of dependencies between variables, depicted below.

    \begin{center}
        \captionof{figure}{Bayesian hierarchical model.}
        \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=0.6cm and 1.2cm,semithick]
            \tikzstyle{every state}=[]

            \node[state] (P) {$\Probmatrix\branchsiteexp$};
            \node[state] (Q) [below right=of P] {$\Submatrix\branchsiteexp$};
            \node[state] (R) [below right=of Q] {$\Mutmatrix$};
            \node[state] (BL) [BLUE, right=of P] {$\branchlength\branchexp$};
            \node[res] (BLH) [RED, right=of BL] {$ \lambda, \tau $};
            \node[state] (f) [right=of Q] {$\Fit\siteexp $};
            \node[state] (Ex) [BLUE, below right=of R] {$\Exchan$};
            \node[state] (Equi) [BLUE, right=of R] {$\Mutequi$};
            \node[state] (Base) [BLUE, above right=of f] {$\Profile\catexp$};
            \node[state] (cat) [BLUE, right=of f] {$\catsite$};
            \node[res] (ExH) [RED, right=of Ex] {$1,1,1,1,1,1$};
            \node[res] (EquiH) [RED, right=of Equi] {$1,1,1,1$};
            \node[state] (baseH) [RED, right=of Base] {$\concentrationProfile, \centerProfile $};
            \node[state] (sb) [BLUE, right=of cat] {$\StickBreaking$};
            \node[state] (sbH) [RED, right=of sb] {$\stickbreakinghyper$};

            \path
            (Q) edge [black] node [above right] {\ref{eq:Probmatrix}} (P)
            (BL) edge [black] node [] {} (P)
            (R) edge [black] node {} (Q)
            (f) edge [black] node [above] {\ref{eq:subrates}} (Q)
            (Ex) edge [black] node [] {} (R)
            (Equi) edge [black] node [below] {\ref{eq:gtr-mutrates}} (R)
            (Base) edge [black] node {} (f)
            (cat) edge [black] node [above] {\ref{eq:sitefitness}} (f)
            (BLH) edge [dashed, BLUE] node [above] {\ref{eq:branchlength}} (BL)
            (ExH) edge [dashed, BLUE] node [above] {\ref{eq:DistribExchan}} (Ex)
            (EquiH) edge [dashed, BLUE] node [above] {\ref{eq:DistribMutequi}} (Equi)
            (baseH) edge [dashed, BLUE] node [above] {\ref{eq:DistribBase}} (Base)
            (sb) edge [dashed, BLUE] node [above] {\ref{eq:DistribMultinomial}} (cat)
            (sbH) edge [dashed, BLUE] node [above] {\ref{eq:DistribStickBreaking},\ref{eq:Beta}} (sb);
        \end{tikzpicture}
    \end{center}

    Nodes of the directed acyclic graph are the variables, and edges are the functions.
    Hyper-parameters are depicted in {\color{RED}{red}} circles, random variables in {\color{BLUE}{blue}} circles, and transformed variables in black.
            {\color{BLUE}{Blue}} dashed line denotes a drawing from a random distribution, and black solid lines denote a function.
    All the nodes pointing toward a given node (upstream) are its dependencies which determine its distribution.
    The other way around, following the arrows in the DAG (downstream), simple {prior} distributions are combined together to form more complex joint {prior} distributions which ultimately define the {prior} distribution of the model.

    \subsection{Nucleotide mutation rates}
    The generalized time-reversible (GTR) nucleotide mutation rate matrix $\Mutmatrix$ is a function of the nucleotide frequencies $\Mutequi$ and the symmetric exchangeability rates $\Exchan$~\cite{tavare_probabilistic_1986}.
    $\Mutequi = (\mutequi_A , \mutequi_C , \mutequi_G , \mutequi_T)$ is the equilibrium base frequency vector, giving the frequency at which each base occurs at each site.
    $\Exchan = \left( \exchan_{AC}, \exchan_{AG}, \exchan_{AT}, \exchan_{CG}, \exchan_{CT}, \exchan_{GT}\right)$ is the vector of exchangeabilities between nucleotides.
    Altogether, the rate matrix is:
    \begin{equation}
        \label{eq:gtr-mutrates}
        \Mutmatrix =
        \begin{blockarray}{ccccc}
            & A & C & G & T \\
            \begin{block}{c(cccc)}
                A & -\mu_A & {\exchan_{AC}\mutequi_C} & {\exchan_{AG}\mutequi_G} & {\exchan_{AT}\mutequi_T} \\
                C & {\exchan_{AC}\mutequi_A} &                        -\mu_C & {\exchan_{CG}\mutequi_G} & {\exchan_{CT}\mutequi_T} \\
                G & {\exchan_{AG}\mutequi_A} & {\exchan_{CG}\mutequi_C} &                        -\mu_G & {\exchan_{GT}\mutequi_T} \\
                T & {\exchan_{AT}\mutequi_A} & {\exchan_{CT}\mutequi_C} & {\exchan_{GT}\mutequi_G} & -\mu_T\\
            \end{block}
        \end{blockarray}
    \end{equation}
    By definition, the sum of the entries in each row of the nucleotide rate matrix $\Mutmatrix$ is equal to $0$, giving the diagonal entries:
    \begin{equation}
        \mu_a = \sum\limits_{ b \neq a, b \in \{A, C, G, T\} } \mutmatrix_{a,b}
    \end{equation}
    The {prior} on the exchangeabilities $\Exchan$ is a uniform Dirichlet distribution of dimension $6$:
    \begin{equation}
        \label{eq:DistribExchan}
        \Exchan \sim \text{Dir}\left( 1,1,1,1,1,1 \right).
    \end{equation}
    The {prior} on the equilibrium base frequencies $\Mutequi$ is a uniform Dirichlet distribution of dimension $4$:
    \begin{equation}
        \label{eq:DistribMutequi}
        \Mutequi \sim \text{Dir}\left( 1,1,1,1 \right)
    \end{equation}
    The general time-reversible nucleotide matrix is normalized with a total flow of $1$:
    \begin{equation}
        \sum\limits_{a \in \{A, C, G, T\}} - \mutequi_a \mutmatrix_{a,a} = 1,
    \end{equation}
    such that we expect $1$ {substitution} per unit of branch length.

    \subsection{Site-specific amino-acid fitness profiles}
    \label{sec:profiles}
    Site-specific amino-acid fitness profiles are assumed i.i.d. from a mixture model, itself endowed with a truncated Dirichlet process prior.
    Specifically, the mixture has $\Ncat$ components ($\Ncat = 30$ by default).
    The prior on component weights~($\StickBreaking$) is modeled using a stick-breaking process, truncated at $\Ncat$ and of parameter $\stickbreakinghyper=1$:
    \begin{align}
        \label{eq:DistribStickBreaking}
        \begin{split}
            & \StickBreaking \sim \text{StickBreaking}\left( \Ncat, \stickbreakinghyper \right)\\
            \iff & \stickbreaking_{\cat} = \stick_{\cat}\cdot \prod _{{\indice=1}}^{{\cat-1}}\left(1-\stick_{\indice}\right),\ \Setcat,
        \end{split}
    \end{align}
    where $\stick_{\cat}$ are i.i.d. from a beta distribution
    \begin{equation}
        \label{eq:Beta}
        \stick_{\cat} \sim \text{Beta}\left( 1, \stickbreakinghyper \right),\ \Setcat.
    \end{equation}
    Of note, the weights decrease geometrically in expectation, at rate $\stickbreakinghyper$, such that lower values of $\stickbreakinghyper$ induce more heterogeneous distributions of weights.

    Each component of the mixture defines a 20-dimensional fitness profile $\Profile\catexp$ (summing to $1$), for $ \Setcat$.
    These fitness profiles are i.i.d. from a Dirichlet of center $\centerProfile=1$ and concentration $\concentrationProfile=1$:
    \begin{equation}
        \label{eq:DistribBase}
        \Profile\catexp \sim \text{Dir}\left( \centerProfile,\ \concentrationProfile \right),\ \Setcat.
    \end{equation}

    Site allocations to the mixture components $\catsite \in \catInterval $, for $\Setsite$ running over the $\Nsite$ sites of the alignment, are i.i.d. multinomial of parameter $\StickBreaking$:
    \begin{align}
        \label{eq:DistribMultinomial}
        \catMultiVar \sim \text{Multinomial}\left( \StickBreaking \right), \\
        \text{ where } \catmultivar_{\cat} = \sum_{\Setsite} \mathbb{1}_{\catsite = \cat}
    \end{align}

    For a given parameter configuration for the mixture, the scaled fitness $\Fit\siteexp$ at site $\site$, are obtained by taking the logarithm of fitness assigned to this site:
    \begin{equation}
        \label{eq:sitefitness}
        \Fit\siteexp = \ln \left( \Profile^{\left( \catsite \right)} \right),\ \Setsite.
    \end{equation}

    \subsection{Branch length}
    The topology of the rooted phylogenetic tree is supposed to be known and is not estimated by the model.
    The branch lengths $\branchlength\branchexp$ are defined as the expected number of {neutral} substitutions per {DNA} site along a branch, each from a Gamma distribution of mean $\lambda=0.1$ and scale $\tau=1$:
    \begin{equation}
        \label{eq:branchlength}
        \branchlength\branchexp \sim \text{Gamma}\left( \lambda, \tau \right).
    \end{equation}

    \subsection{Codon substitution rates}

    The mutation rate between codons $\ci$ and $\cj$, denoted $\mu_{\itoj}$ depends on the underlying nucleotide change between the codons.
    First, if codons $\ci$ and $\cj$ are not nearest-neighbours, $\mu_{\itoj}$ is equal to $0$.
    Second, if codons $\ci$ and $\cj$ are only one mutation away, $\mu_{\itoj}$ is given by the underlying nucleotide relative rate~(${\mutmatrix_{\itoj}}$).

    For a given site $\site$, the {codon} {substitution} rate matrix $\Submatrix\siteexp$ is given by:
    \begin{equation}
        \label{eq:subrates}
        \begin{dcases}
            \submatrix\siteexp_{\itoj} = 0\text{ if codons $\ci$ and $\cj$ are not nearest-neighbors,} \\
            \submatrix\siteexp_{\itoj} = \mu_{\itoj}\text{ if codons $\ci$ and $\cj$ are {synonymous},} \\
            \submatrix\siteexp_{\itoj} = \mu_{\itoj} \dfrac{ \fitj\siteexp - \fiti\siteexp}{{1 - \e^{\fiti\siteexp - \fitj\siteexp} }} \text{ if $\ci$ and $\cj$ are non-synonymous,}\\
            \submatrix\siteexp_{\ci, \ci} = - \sum\limits_{ \cj \neq \ci, \cj=1}^{61} \submatrix\siteexp_{\itoj}.
        \end{dcases}
    \end{equation}
    Together, the probability of transition between codons for a given branch $\branch$ and site $\site$ is:
    \begin{equation}
        \label{eq:Probmatrix}
        \Probmatrix\branchsiteexp = \e^{\branchlength\branchexp \Submatrix\siteexp},
    \end{equation}
    which are the matrices necessary to compute the {likelihood} of the data ($\data$) given the parameters of the model using the pruning algorithm.

    \subsection{Bayesian implementation}
    \label{sec:Bayesian}
    Bayesian inference was conducted using Markov Chain Monte Carlo (MCMC).
    Most phylogenetic {MCMC} samplers target the distribution over the model parameters given the sequence alignment, which means that they have to repeatedly invoke the pruning algorithm to recalculate the {likelihood} which is most often the limiting step of the {MCMC}.
    An alternative, which is used here, is to do the {MCMC} conditionally on the detailed {substitution} history $\subhistory$, thus doing the {MCMC} over the augmented configuration~($\subhistory$, $\data$), under the target distribution obtained by combining the mapping-based {likelihood} with the {prior} over model parameters.
    The key idea that makes this strategy efficient is that the mapping-based {likelihood} depends on compact summary statistics of $\subhistory$, leading to very fast evaluation of the {likelihood}.
    On the other hand, this requires to implement more complex {MCMC} procedures that have to alternate between:
    \begin{enumerate}
        \item sampling $\subhistory$ conditionally on the data and the current parameter configuration.
        \item re-sampling the parameters conditionally on $\subhistory$.
    \end{enumerate}

    To implement the mapping-based {MCMC} sampling strategy, we first sample the detailed {substitution} history $\subhistory$ for all sites along the tree.
    Several methods exist for doing this~\cite{nielsen_mapping_2002,rodrigue_uniformization_2008}, which are used here in combination (first trying the accept-reject method of \textcite{nielsen_mapping_2002}, then switching to the uniformization approach of \textcite{rodrigue_uniformization_2008} if the first round has failed).

    Then, we write down the probability of $\subhistory$ given the parameters, and finally, we collect all factors that depend on some parameter of interest and make some simplifications.
    This ultimately leads to relatively compact sufficient statistics allowing for fast numerical evaluation of the likelihood~\cite{irvahn_phylogenetic_2014,davydov_state_2017}.

    \subsection{BayesCode software}\label{subsec:bayescode}
    In \textit{BayesCode} (\href{https://github.com/ThibaultLatrille/BayesCode}{github.com/ThibaultLatrille/BayesCode}, v1.3.1), we ran the mutation-selection codon models \textit{mutselomega} for 2000 points of MCMC with the options:
    \begin{scriptsize}
        \begin{verbatim}
 mutselomega ---omegashift 0.0 --ncat 30 -a my_alignment.phy -t my_tree.newick -u 2000 my_genename
        \end{verbatim}
    \end{scriptsize}
    The collection of site-specific fitness profiles ($\UniDimArray{F^{(i)}}, \forall i$) are then obtained by running \textit{readmutselomega}, reading 1000 points of MCMC (first 1000 are considered as burn-in) with the options:
    \begin{scriptsize}
        \begin{verbatim}
 readmutselomega --every 1 --until 2000 --burnin 1000 --ss my_genename
        \end{verbatim}
    \end{scriptsize}
    The gene-specific $4 \times 4$ nucleotide mutation rate matrix ($\UniDimArray{\mu}$) is also obtained by running \textit{readmutselomega}, reading 1000 points of MCMC (first 1000 are considered as burn-in) with the options:
    \begin{scriptsize}
        \begin{verbatim}
 readmutselomega --every 1 --until 2000 --burnin 1000 --nuc my_genename
        \end{verbatim}
    \end{scriptsize}

@@

    \subsection{Example sites}\label{subsec:example-sites}

    We extracted protein-coding DNA alignments across mammals centered around the codon site (two flanking codons in white background) for which the beneficial non-adaptive mutations ($\SphyBen$) have been detected in \textit{Chlorocebus sabaeus}.
    We also show the translated amino acids of this region.
    For instance, in DNA alignment of gene SELE shown below, the nucleotide at site 1722 has mutated (from T to C) at the basis of Simiiformes (monkeys and apes), modifying the corresponding amino acid from Serine to Proline, but has been subsequently reverted in the branch of \textit{Chlorocebus sabaeus}.
    However, other substitutions classified as $\SphyBen$ cannot be clearly interpreted as reversions \textit{sensu stricto} along the terminal branch of \textit{Chlorocebus sabaeus}.
    Indeed, we acknowledge that on a fixed fitness landscape, a deleterious mutation can be compensated by transitions to other fitter amino-acids, and not necessarily the ancestral one.

    In the figures below, the first line is the name of the gene, the second line the position of the mutation in the OrthoMam protein-coding DNA alignment, and the arrow in the third line corresponds to the nucleotide position of the mutation.
    Finally, for reproducibility, we generated the fasta alignment for all $\SphyBen$ mutations that we detected, either in the terminal lineage or in segregating polymorphism.
    These fasta files are available at Zenodo (\url{https://doi.org/10.5281/zenodo.7878953}), across all populations, under the compressed archive \textit{alignment\_focused\_around\_non\_adaptive\_mutations.zip} and we provide a python script to obtain the figures shown below given these alignments.

    \newpage
    \begin{center}
        \captionof{figure}{Examples sites of $\SphyBen$ mutations in \textit{Chlorocebus sabaeus} (reversions).}.
        \includegraphics[width=0.95\linewidth, page=1]{artworks/Alignment.Chlorocebus_sabaeus.Saint_Kitts-0.pdf}
    \end{center}

    \newpage
    \begin{center}
        \captionof{figure}{Examples sites of $\SphyBen$ mutations in \textit{Chlorocebus sabaeus}.}.
        \includegraphics[width=0.95\linewidth, page=1]{artworks/Alignment.Chlorocebus_sabaeus.Saint_Kitts-1.pdf}
    \end{center}

    \newpage
    \subsection{Selection along the terminal branches}\label{subsec:summary-table-mutsel}

    \subsubsection{Probability of mutations and substitutions to be $\SphyDel$, $\SphyNeu$ or $\SphyBen$}
    \begin{center}
        \captionof{table}{Probability of mutations and substitutions to be $\SphyDel$, $\SphyNeu$ or $\SphyBen$.}
        \scriptsize
        \begin{longtable*}{|l|l|r|r|r|r|r|r|}
            \toprule
            Population &             Species & $\proba[ \SphyDel ]$ & $\proba[ \SphyNeu ]$ & $\proba[ \SphyBen ]$ & $\proba_{div}[ \SphyDel ]$ & $\proba_{div}[ \SphyNeu ]$ & $\proba_{div}[ \SphyBen ]$ \\
            \midrule
            \endhead
            \midrule
            \multicolumn{8}{r}{{Continued on next page}} \\
            \midrule
            \endfoot

@@

        \item $\proba [ \SphyDel ]$ (eq.~5) is the probability for a new mutation to be a deleterious.
        These mutations have a selection coefficient predicted at the phylogenetic-scale lower than -1, thus toward a less fit amino-acid.
        \item $\proba [ \SphyNeu ]$ (eq.~5) is the probability for a new mutation to be a nearly-neutral.
        These mutations have a selection coefficient predicted at the phylogenetic-scale between -1 and 1.
        \item $\proba [ \SphyBen ]$ (eq.~5) is the probability for a new mutation to be a  non-adaptive beneficial mutation.
        \item $\proba_{div}[\SphyDel]$ is the proportion of substitutions in the terminal branch that are $\SphyDel$.
        \item $\proba_{div}[\SphyNeu]$ is the proportion of substitutions in the terminal branch that are $\SphyNeu$.

@@

@@

    \subsubsection{$\dnds$ for $\SphyDel$, $\SphyNeu$ or $\SphyBen$}
    \begin{center}
        \captionof{table}{$\dnds$ for $\SphyDel$, $\SphyNeu$ or $\SphyBen$.}
        \scriptsize
        \begin{longtable*}{|l|r|r|r|r|}
            \toprule
            Species & $\dnds $ & $\dn ( \SphyDel ) / \ds$ & $\dn ( \SphyNeu ) / \ds$ & $\dn ( \SphyBen ) / \ds$ \\
            \midrule
            \endhead
            \midrule
            \multicolumn{5}{r}{{Continued on next page}} \\
            \midrule
            \endfoot

            \bottomrule
            \endlastfoot
            \rowcolor{LIGHTGREY} Equus caballus & $ 0.129$ & $ 0.065$ & $ 0.832$ & $ 1.267$ \\
            Bos taurus & [$ 0.114$, $ 0.116$] & [$ 0.063$, $ 0.064$] & [$ 0.629$, $ 0.638$] & [$ 1.280$, $ 1.328$] \\
            \rowcolor{LIGHTGREY} Capra hircus & [$ 0.108$, $ 0.109$] & [$ 0.058$, $ 0.058$] & [$ 0.631$, $ 0.636$] & [$ 1.169$, $ 1.183$] \\
            Ovis aries & [$ 0.127$, $ 0.129$] & [$ 0.078$, $ 0.080$] & [$ 0.619$, $ 0.621$] & [$ 1.201$, $ 1.217$] \\
            \rowcolor{LIGHTGREY} Chlorocebus sabaeus & [$ 0.118$, $ 0.119$] & [$ 0.062$, $ 0.062$] & [$ 0.713$, $ 0.720$] & [$ 1.521$, $ 1.577$] \\
            Homo sapiens & [$ 0.170$, $ 0.170$] & [$ 0.103$, $ 0.103$] & [$ 0.884$, $ 0.888$] & [$ 1.733$, $ 1.745$] \\
        \end{longtable*}
    \end{center}
    \begin{itemize}
        \item $\dnds$ (eq.~6) is the ratio of non-synonymous over synonymous substitutions estimated for all the non-synonymous substitutions in the terminal branch.
        \item $\dn(\SphyDel) / \ds$ (eq.~6) is the ratio of non-synonymous over synonymous substitutions, when restricted to non-synonymous substitutions in the terminal branch that are $\SphyDel$.
        \item $\dn(\SphyNeu) / \ds$ (eq.~6) is the ratio of non-synonymous over synonymous substitutions, when restricted to non-synonymous substitutions in the terminal branch that are $\SphyNeu$.
        \item $\dn(\SphyBen) / \ds$ (eq.~6) is the ratio of non-synonymous over synonymous substitutions, when restricted to non-synonymous substitutions in the terminal branch that are $\SphyBen$.
    \end{itemize}
    SNP are considered fixed in the populations if all sampled individuals are homozygous for the derived allele, in which case we consider this SNP not to be segregating but as fixed as a substitution.
    This effect results in $\dn / \ds$ that can change across populations, denoted as a range per species.
    $\dn(\SphyBen) / \ds$ values observed are above 1 and are consistent with slightly advantageous mutations.\\

    Theoretically, $\omega = \dn / \ds$ can be related to the underlying scaled selection coefficient ($S$) with the relation $\omega = S /(1-\exp(-S))$ as in \cite[eq. 3]{nielsen_estimating_2003}.
    In our experiments our observed $\dn(\SphyBen) / \ds$ is on the range $1.169$ (\textit{C. hircus}) to 1.745 (\textit{H. sapiens}), translating into average $S$ of $\approx0.32$ to $\approx1.24$, so indeed only slightly advantageous.\\

    Of note, observing $\dn(\SphyBen) / \ds >1$ is an important check that $\SphyBen$ mutations are indeed positively selected with an increased substitution rate.
    Since $\dn(\SphyNeu) / \ds$ is close to 1 for the predicted nearly-neutral mutations, this means that the selection coefficients predicted at the mutation-selection balance are good proxies of selection, then observing $\dn(\SphyBen) / \ds > 1$ is evidence of positive selection of predicted non-adaptive beneficial mutations.

    \newpage
    \subsubsection{$\dnds$ over-estimation due to non-adaptive beneficial mutations}
    \begin{center}
        \captionof{table}{$\dnds$ over-estimation due to non-adaptive beneficial mutations.}
        \scriptsize
        \begin{longtable*}{|l|r|r|r|}
            \toprule
            Species & $\dnds $ & $\dn(\Sphy < 1) / \ds$ & $\delta(\dnds )$ \\
            \midrule
            \endhead
            \midrule
            \multicolumn{4}{r}{{Continued on next page}} \\
            \midrule
            \endfoot
